---
title: "gstat exercise"
output: html_notebook
---

from: 
https://cran.r-project.org/web/packages/gstat/vignettes/gstat.pdf
https://github.com/r-spatial/gstat/blob/master/vignettes/gstat.Rnw

###3 Spatial data frames
As an example, we will look at the meuse data set, which is a regular data frame
that comes with package gstat (remove the 88 from the colour strings to make
a plot without alpha transparency on windows or X11):

```{r}
library(sp)
data(meuse)
class(meuse)
names(meuse)
coordinates(meuse) = ~x+y
class(meuse)
summary(meuse)
coordinates(meuse)[1:5,]
bubble(meuse, "zinc", 
	col=c("#00ff0088", "#00ff0088"), main = "zinc concentrations (ppm)")
```

###Spatial data on a regular grid

```{r}
data(meuse.grid)
summary(meuse.grid)
class(meuse.grid)
coordinates(meuse.grid) = ~x+y
class(meuse.grid)
gridded(meuse.grid) = TRUE
class(meuse.grid)
image(meuse.grid["dist"])
title("distance to river (red = 0)")
library(gstat)
zinc.idw = idw(zinc~1, meuse, meuse.grid)
class(zinc.idw)
spplot(zinc.idw["var1.pred"], main = "zinc inverse distance weighted interpolations")
```

If you compare the bubble plot of zinc measurements with the map with
distances to the river, it becomes evident that the larger concentrations
are measured at locations close to the river. This relationship can be
linearized by log-transforming the zinc concentrations, and taking the
square root of distance to the river:

```{r}
plot(log(zinc)~sqrt(dist), meuse)
abline(lm(log(zinc)~sqrt(dist), meuse))
```






###5 Variograms
Variograms are calculated using the function variogram, which takes a formula
as its first argument: log(zinc)~1 means that we assume a constant trend for
the variable log(zinc).

```{r}
library(gstat)
lzn.vgm = variogram(log(zinc)~1, meuse)
lzn.vgm
lzn.fit = fit.variogram(lzn.vgm, model = vgm(1, "Sph", 900, 1))
lzn.fit
plot(lzn.vgm, lzn.fit)
```


Instead of the constant mean, denoted by \verb|~1|, we can specify a
mean function, e.g. using \verb|~sqrt(dist)| as a predictor variable:

```{r}
lznr.vgm = variogram(log(zinc)~sqrt(dist), meuse)
lznr.fit = fit.variogram(lznr.vgm, model = vgm(1, "Exp", 300, 1))
lznr.fit
plot(lznr.vgm, lznr.fit)
```


In this case, the variogram of residuals with respect to a fitted mean
function are shown. Residuals were calculated using ordinary least
squares.

###Kriging

```{r}
lzn.kriged = krige(formula = log(zinc)~1, meuse, meuse.grid, model = lzn.fit)
spplot(lzn.kriged["var1.pred"])
spplot(lzn.kriged)
```












